Variational Inference for Inverse Reinforcement Learning with Gaussian Processes

### 2019-03-21 mpaper/paper.tex
The proof is now correct, complete, although could be cleaned up a bit

\section{Theoretical Justification} \label{sec:proof}

@@

% TODO: need some editing
\begin{lemma} \label{lemma:bound1}
  Let $i \in \{ 0, \dots, d \}$ and $\epsilon > 0$ be arbitrary. Furthermore,
  let $c : \mathbb{R}^{|\mathcal{S}|} \times \mathbb{R}^m \to (\lambda_i -
  \epsilon, \lambda_i + \epsilon) \subset \mathbb{R}$ be a function with a
  codomain arbitrarily close to $\lambda_i$. Then
  \[
    \left. \frac{\partial q(\mathbf{r} \mid \mathbf{u})}{\partial \lambda_i}
    \right|_{\lambda_i = c(\mathbf{r}, \mathbf{u})}
  \]
  has upper and lower bounds of the form $q(\mathbf{r} \mid
  \mathbf{u})d(\mathbf{u})$, where $d(\mathbf{u}) \in \mathbb{R}_2[\mathbf{u}]$.
\end{lemma}
\begin{proof}
  Remember that
  \begin{align*}
    \frac{\partial q(\mathbf{r} \mid \mathbf{u})}{\partial \lambda_i} &= \frac{1}{2}q(\mathbf{r} \mid \mathbf{u}) (|\bm\Gamma|^{-1} \tr(\mathbf{R} \adj(\bm\Gamma)) \\
                                                                      &- (\mathbf{r} - \mathbf{Su})^\intercal\bm\Gamma^{-1}\mathbf{R}\bm\Gamma^{-1}(\mathbf{r} - \mathbf{Su})).
  \end{align*}
  by Lemma \ref{lemma:derivatives}. Let $\mathbf{K}$ be any covariance
  matrix and
  \[
    \mathbf{A} = \frac{1}{\lambda_0} \mathbf{K}.
  \]

  First, we will show that
  \[
    \mathbf{K}|_{\lambda_i = c(\mathbf{r}, \mathbf{u})} \to 0 \quad \text{and}
    \quad \left. \frac{\mathbf{\partial K}}{\partial \lambda_i}
    \right|_{\lambda_i = c(\mathbf{r}, \mathbf{u})} \to 0
  \]
  as $\epsilon \to 0$.
  We can easily establish constant upper and lower bounds on
  $\mathbf{K}|_{\lambda_i = c(\mathbf{r}, \mathbf{u})}$ by the boundedness of
  $c$ and continuity of the covariance function. Similar reasoning combined with
  the expressions for derivatives of covariance matrices in Lemma
  \ref{lemma:derivatives} gives constant upper and lower bounds on the elements
  of
  \[
    \left. \frac{\mathbf{\partial K}}{\partial \lambda_i} \right|_{\lambda_i =
      c(\mathbf{r}, \mathbf{u})}
  \]
  as well.

  % Inverse of a covariance matrix
  Now, we will show that $\mathbf{K}^{-1}|_{\lambda_i = c(\mathbf{r},
    \mathbf{u})}$ exists and
  \[
    \lim_{\epsilon \to 0} \mathbf{K}^{-1}|_{\lambda_i = c(\mathbf{r},
      \mathbf{u})} = \mathbf{K}.
  \]
  If $i = 0$, then $\mathbf{K}|_{\lambda_i = c(\mathbf{r}, \mathbf{u})} =
  c(\mathbf{r}, \mathbf{u})\mathbf{A}$. Therefore\footnote{Note that since
    $\lambda_0 \ne 0$, $c(\mathbf{r}, \mathbf{u}) \ne 0$ for small enough
    $\epsilon$.},
  \begin{align*}
    \mathbf{K}^{-1}|_{\lambda_i = c(\mathbf{r}, \mathbf{u})} &=
                                                               \frac{1}{c(\mathbf{r}, \mathbf{u})}\mathbf{A}^{-1} \\
                                                             &\to \frac{1}{\lambda_0}\mathbf{A}^{-1} = \frac{1}{\lambda_0} \left( \frac{1}{\lambda_0}\mathbf{K} \right)^{-1} = \mathbf{K}^{-1}
  \end{align*}
  as $\epsilon \to 0$.

  We will now show the same result for $\lambda_i > 0$. Let
  \[
    S = \sum_{n \in \{ 1, \dots, d \} \setminus \{ i \}} \frac{\lambda_n}{2}
    (x_{j,n} - x_{k,n})^2 + \mathbbm{1}[j \ne k]\sigma^2 \lambda_n
  \]
  and $\delta = c(\mathbf{r}, \mathbf{u}) - \lambda_i$ so that $c(\mathbf{r},
  \mathbf{u}) = \lambda_i + \delta$, and $\lim_{\epsilon \to 0} \delta = 0$.
  Then,
  \begin{align*}
    &k(\mathbf{x}_j, \mathbf{x}_k)|_{\lambda_i = c(\mathbf{r}, \mathbf{u})} = \!\begin{multlined}[t]
      \lambda_0 \exp \left( \vphantom{\frac{1}{2}} \right. -\frac{1}{2} c(\mathbf{r}, \mathbf{u})(x_{j,i} - x_{k,i})^2\\
      - \mathbbm{1}[j \ne k]\sigma^2c(\mathbf{r}, \mathbf{u}) - S \left. \vphantom{\frac{1}{2}} \right)
    \end{multlined} \\
    &= \!\begin{multlined}[t]
      \lambda_0 \exp \left( \vphantom{\frac{1}{2}} \right. -\frac{1}{2} (\lambda_i + \delta)(x_{j,i} - x_{k,i})^2 \\
      - \mathbbm{1}[j \ne k]\sigma^2(\lambda_i + \delta) - S \left. \vphantom{\frac{1}{2}} \right)
    \end{multlined} \\
    &= \!\begin{multlined}[t]
      \lambda_0 \exp \left( \vphantom{\frac{1}{2}} \right. -\frac{1}{2}(\mathbf{x}_j - \mathbf{x}_k)^\intercal \bm\Lambda (\mathbf{x}_j - \mathbf{x}_k) - \mathbbm{1}[j \ne k]\sigma^2\tr(\bm\Lambda) \\
      - \frac{\delta}{2}(x_{j,i} - x_{k,i})^2 - \mathbbm{1}[j \ne k]\sigma^2\delta \left. \vphantom{\frac{1}{2}} \right)
    \end{multlined} \\
    &= k(\mathbf{x}_j, \mathbf{x}_k)\exp \left( \vphantom{\frac{1}{2}} \right. -\frac{\delta}{2}(x_{j,i} - x_{k,i})^2 - \mathbbm{1}[j \ne k]\sigma^2\delta \left. \vphantom{\frac{1}{2}} \right) \\
    &= \!\begin{multlined}[t]
      k(\mathbf{x}_j, \mathbf{x}_k) + k(\mathbf{x}_j, \mathbf{x}_k) \left( \exp
        \left( -\frac{\delta}{2}(x_{j,i} - x_{k,i})^2 \right. \right. \\
      -\mathbbm{1}[j \ne k]\sigma^2\delta \left. \left. \vphantom{\frac{1}{2}} \right) - 1 \right)
    \end{multlined}
  \end{align*}
  Hence, we can express $\mathbf{K}|_{\lambda_i = c(\mathbf{r}, \mathbf{u})}$ as
  $\mathbf{K}|_{\lambda_i = c(\mathbf{r}, \mathbf{u})} = \mathbf{K} +
  \mathbf{E}$, where $\mathbf{E}$ is defined as
  \[
    E_{j,k} = k(\mathbf{x}_j, \mathbf{x}_k) \left( \exp \left(
        -\frac{\delta}{2}(x_{j,i} - x_{k,i})^2 \right. \right. -\mathbbm{1}[j
    \ne k]\sigma^2\delta \left. \left. \vphantom{\frac{1}{2}} \right) - 1
    \right).
  \]
  By this definition,
  \[
    \lim_{\epsilon \to 0} E_{j,k} = 0.
  \]
  Then, since $\mathbf{K}$ is invertible, Lemma \ref{lemma:perturbation} shows
  the existence of $\mathbf{K}^{-1}_{\lambda_i = c(\mathbf{r}, \mathbf{u})}$ and
  gives upper and lower bounds on all of its elements.

  This is enough to prove constant upper and lower bounds on
  $\mathbf{S}$, $\bm\Gamma$, and $\mathbf{R}$ (all with $\lambda_i$ replaced
  with $c(\mathbf{r}, \mathbf{u})$), which means that $(\mathbf{r} -
  \mathbf{Su})^\intercal\bm\Gamma^{-1}\mathbf{R}\bm\Gamma^{-1}(\mathbf{r} -
  \mathbf{Su})|_{\lambda_i = c(\mathbf{r}, \mathbf{u})}$ has upper and lower
  bounds in $\mathbb{R}_2[\mathbf{u}]$.

  Since
  \[
    \lim_{\epsilon \to 0} \bm\Gamma|_{\lambda_i = c(\mathbf{r}, \mathbf{u})} =
    \bm\Gamma,
  \]
  we also have that
  \[
    \lim_{\epsilon \to 0} \det(\bm\Gamma)|_{\lambda_i = c(\mathbf{r}, \mathbf{u})} =
    \det(\bm\Gamma).
  \]
  Assuming that
  $\bm\Gamma$ is invertible so that $q(\mathbf{r} \mid \mathbf{u})$ exists,
  \[
    \det(\bm\Gamma)|_{\lambda_i = c(\mathbf{r}, \mathbf{u})} \ne 0
  \]
  for small enough $\epsilon$, and, thus, $\det(\bm\Gamma)^{-1}|_{\lambda_i =
    c(\mathbf{r}, \mathbf{u})}$ exists and has constant bounds.

@@

  Since we already know that $\bm\Gamma^{-1}|_{\lambda_i = c(\mathbf{r},
    \mathbf{u})}$ and $\det(\bm\Gamma)|_{\lambda_i = c(\mathbf{r}, \mathbf{u})}$
  are bounded, so is $\adj(\bm\Gamma) =
  \det(\bm\Gamma)\bm\Gamma^{-1}|_{\lambda_i = c(\mathbf{r}, \mathbf{u})}$. Thus,
  we have constant bounds on
  $|\bm\Gamma|^{-1}\tr(\mathbf{R}\adj(\bm\Gamma))|_{\lambda_i = c(\mathbf{r},
    \mathbf{u})}$, which means that
  \[
    \left. \frac{\partial q(\mathbf{r} \mid \mathbf{u})}{\partial \lambda_i}
    \right|_{\lambda_i = c(\mathbf{r}, \mathbf{u})}
  \]
  has the required bounds.
\end{proof}
% TOOD: epsilon should not be treated as a constant. Revise the whole theory
% section.

>>>>

\section{Theoretical Justification} \label{sec:proof}
% TODO: Possible improvements:
% 1. Use limits more often.
% 2. Perturbation lemma can be replaced by continuity of A |-> A^-1.

@@

% TODO: too long, lemma could be replaced by the continuity argument (and arrow
% removed from the figure), needs more explanations. Otherwise quite tolerable.
\begin{lemma} \label{lemma:bound1}
  Let $i \in \{ 0, \dots, d \}$ and $\epsilon > 0$ be arbitrary. Furthermore,
  let $c : \mathbb{R}^{|\mathcal{S}|} \times \mathbb{R}^m \to (\lambda_i -
  \epsilon, \lambda_i + \epsilon) \subset \mathbb{R}$ be a function with a
  codomain arbitrarily close to $\lambda_i$. Then
  \[
    \left. \frac{\partial q(\mathbf{r} \mid \mathbf{u})}{\partial \lambda_i}
    \right|_{\lambda_i = c(\mathbf{r}, \mathbf{u})}
  \]
  has upper and lower bounds of the form $q(\mathbf{r} \mid
  \mathbf{u})d(\mathbf{u})$, where $d(\mathbf{u}) \in \mathbb{R}_2[\mathbf{u}]$.
\end{lemma}
\begin{proof}
  Remember that
  \begin{align*}
    \frac{\partial q(\mathbf{r} \mid \mathbf{u})}{\partial \lambda_i} &= \frac{1}{2}q(\mathbf{r} \mid \mathbf{u}) (|\bm\Gamma|^{-1} \tr(\mathbf{R} \adj(\bm\Gamma)) \\
                                                                      &- (\mathbf{r} - \mathbf{Su})^\intercal\bm\Gamma^{-1}\mathbf{R}\bm\Gamma^{-1}(\mathbf{r} - \mathbf{Su})).
  \end{align*}
  by Lemma \ref{lemma:derivatives}. Let $\mathbf{K}$ be any covariance
  matrix and
  \[
    \mathbf{A} = \frac{1}{\lambda_0} \mathbf{K}.
  \]

  % K and its derivative
  First, we can easily deduce that
  \begin{equation} \label{eq:convergence1}
    \mathbf{K}|_{\lambda_i = c(\mathbf{r}, \mathbf{u})} \to \mathbf{K}
  \end{equation}
  and
  \begin{equation} \label{eq:convergence2}
    \left. \frac{\partial \mathbf{K}}{\partial \lambda_i} \right|_{\lambda_i = c(\mathbf{r}, \mathbf{u})} \to \frac{\partial \mathbf{K}}{\partial \lambda_i}
  \end{equation}
  as $\epsilon \to 0$. For \eqref{eq:convergence1}, the result is obvious if $i
  = 0$. Otherwise, it follows from the continuity of the exponential function.
  For \eqref{eq:convergence2}, observe that if $i=0$, then
  \[
    \left. \frac{\partial \mathbf{K}}{\partial \lambda_0} \right|_{\lambda_0 =
      c(\mathbf{r}, \mathbf{u})}
  \]
  as an expression has no $\lambda_0$ in it, so
  \[
    \left. \frac{\partial \mathbf{K}}{\partial \lambda_0} \right|_{\lambda_0 =
      c(\mathbf{r}, \mathbf{u})} = \frac{\partial \mathbf{K}}{\partial
      \lambda_0}.
  \]
  Finally, if $i>0$, then each element of 
  \[
    \left. \frac{\partial \mathbf{K}}{\partial \lambda_i}
    \right|_{\lambda_i = c(\mathbf{r}, \mathbf{u})}
  \]
  is a constant multiple of the corresponding element of
  \[
    \mathbf{K}|_{\lambda_i = c(\mathbf{r}, \mathbf{u})},
  \]
  so the same reasoning as for \eqref{eq:convergence1} applies.

  % Inverse of a covariance matrix
  Now, we will show that $\mathbf{K}^{-1}|_{\lambda_i = c(\mathbf{r},
    \mathbf{u})}$ exists and
  \[
    \lim_{\epsilon \to 0} \mathbf{K}^{-1}|_{\lambda_i = c(\mathbf{r},
      \mathbf{u})} = \mathbf{K}.
  \]
  If $i = 0$, then $\mathbf{K}|_{\lambda_i = c(\mathbf{r}, \mathbf{u})} =
  c(\mathbf{r}, \mathbf{u})\mathbf{A}$. Therefore\footnote{Note that since
    $\lambda_0 \ne 0$, $c(\mathbf{r}, \mathbf{u}) \ne 0$ for small enough
    $\epsilon$.},
  \begin{align*}
    \mathbf{K}^{-1}|_{\lambda_i = c(\mathbf{r}, \mathbf{u})} &=
                                                               \frac{1}{c(\mathbf{r}, \mathbf{u})}\mathbf{A}^{-1} \\
                                                             &\to \frac{1}{\lambda_0}\mathbf{A}^{-1} = \frac{1}{\lambda_0} \left( \frac{1}{\lambda_0}\mathbf{K} \right)^{-1} = \mathbf{K}^{-1}
  \end{align*}
  as $\epsilon \to 0$. For $i > 0$, let
  \[
    S = \sum_{n \in \{ 1, \dots, d \} \setminus \{ i \}} \frac{\lambda_n}{2}
    (x_{j,n} - x_{k,n})^2 + \mathbbm{1}[j \ne k]\sigma^2 \lambda_n
  \]
  and $\delta = c(\mathbf{r}, \mathbf{u}) - \lambda_i$ so that $c(\mathbf{r},
  \mathbf{u}) = \lambda_i + \delta$, and $\lim_{\epsilon \to 0} \delta = 0$.
  Then,
  \begin{align*}
    &k(\mathbf{x}_j, \mathbf{x}_k)|_{\lambda_i = c(\mathbf{r}, \mathbf{u})} = \!\begin{multlined}[t]
      \lambda_0 \exp \left( \vphantom{\frac{1}{2}} \right. -\frac{1}{2} c(\mathbf{r}, \mathbf{u})(x_{j,i} - x_{k,i})^2\\
      - \mathbbm{1}[j \ne k]\sigma^2c(\mathbf{r}, \mathbf{u}) - S \left. \vphantom{\frac{1}{2}} \right)
    \end{multlined} \\
    &= \!\begin{multlined}[t]
      \lambda_0 \exp \left( \vphantom{\frac{1}{2}} \right. -\frac{1}{2} (\lambda_i + \delta)(x_{j,i} - x_{k,i})^2 \\
      - \mathbbm{1}[j \ne k]\sigma^2(\lambda_i + \delta) - S \left. \vphantom{\frac{1}{2}} \right)
    \end{multlined} \\
    &= \!\begin{multlined}[t]
      \lambda_0 \exp \left( \vphantom{\frac{1}{2}} \right. -\frac{1}{2}(\mathbf{x}_j - \mathbf{x}_k)^\intercal \bm\Lambda (\mathbf{x}_j - \mathbf{x}_k) - \mathbbm{1}[j \ne k]\sigma^2\tr(\bm\Lambda) \\
      - \frac{\delta}{2}(x_{j,i} - x_{k,i})^2 - \mathbbm{1}[j \ne k]\sigma^2\delta \left. \vphantom{\frac{1}{2}} \right)
    \end{multlined} \\
    &= k(\mathbf{x}_j, \mathbf{x}_k)\exp \left( \vphantom{\frac{1}{2}} \right. -\frac{\delta}{2}(x_{j,i} - x_{k,i})^2 - \mathbbm{1}[j \ne k]\sigma^2\delta \left. \vphantom{\frac{1}{2}} \right) \\
    &= \!\begin{multlined}[t]
      k(\mathbf{x}_j, \mathbf{x}_k) + k(\mathbf{x}_j, \mathbf{x}_k) \left( \exp
        \left( -\frac{\delta}{2}(x_{j,i} - x_{k,i})^2 \right. \right. \\
      -\mathbbm{1}[j \ne k]\sigma^2\delta \left. \left. \vphantom{\frac{1}{2}} \right) - 1 \right).
    \end{multlined}
  \end{align*}
  Thus, we can express $\mathbf{K}|_{\lambda_i = c(\mathbf{r}, \mathbf{u})}$ as
  $\mathbf{K}|_{\lambda_i = c(\mathbf{r}, \mathbf{u})} = \mathbf{K} +
  \mathbf{E}$, where $\mathbf{E}$ is defined by
  \[
    E_{j,k} = k(\mathbf{x}_j, \mathbf{x}_k) \left( \exp \left(
        -\frac{\delta}{2}(x_{j,i} - x_{k,i})^2 \right. \right. -\mathbbm{1}[j
    \ne k]\sigma^2\delta \left. \left. \vphantom{\frac{1}{2}} \right) - 1
    \right).
  \]
  By this definition,
  \[
    \lim_{\epsilon \to 0} E_{j,k} = 0.
  \]
  Then, since $\mathbf{K}$ is invertible,
  \[
    \mathbf{K}^{-1}_{\lambda_i = c(\mathbf{r}, \mathbf{u})} \to \mathbf{K}^{-1}
  \]
  by continuity of $\mathbf{A} \mapsto \mathbf{A}^{-1}$.

  This is enough to prove constant upper and lower bounds on
  $\mathbf{S}$, $\bm\Gamma$, and $\mathbf{R}$ (all with $\lambda_i$ replaced
  with $c(\mathbf{r}, \mathbf{u})$), which means that $(\mathbf{r} -
  \mathbf{Su})^\intercal\bm\Gamma^{-1}\mathbf{R}\bm\Gamma^{-1}(\mathbf{r} -
  \mathbf{Su})|_{\lambda_i = c(\mathbf{r}, \mathbf{u})}$ has upper and lower
  bounds in $\mathbb{R}_2[\mathbf{u}]$. Furthermore, having convergence results
  for arbitrary covariance matrices and their inverses means that
  \[
    \lim_{\epsilon \to 0} \bm\Gamma|_{\lambda_i = c(\mathbf{r}, \mathbf{u})} =
    \bm\Gamma.
  \]
  Hence,
  \[
    \lim_{\epsilon \to 0} \det(\bm\Gamma)|_{\lambda_i = c(\mathbf{r}, \mathbf{u})} =
    \det(\bm\Gamma).
  \]
  Assuming that
  $\bm\Gamma$ is invertible so that $q(\mathbf{r} \mid \mathbf{u})$ exists,
  \[
    \det(\bm\Gamma)|_{\lambda_i = c(\mathbf{r}, \mathbf{u})} \ne 0
  \]
  for small enough $\epsilon$, and, thus, $\det(\bm\Gamma)^{-1}|_{\lambda_i =
    c(\mathbf{r}, \mathbf{u})}$ exists and is bounded.

@@

  Since we already know that $\bm\Gamma^{-1}|_{\lambda_i = c(\mathbf{r},
    \mathbf{u})}$ and $\det(\bm\Gamma)|_{\lambda_i = c(\mathbf{r}, \mathbf{u})}$
  are bounded, so is $\adj(\bm\Gamma) =
  \det(\bm\Gamma)\bm\Gamma^{-1}|_{\lambda_i = c(\mathbf{r}, \mathbf{u})}$. Thus,
  we have constant bounds on
  $|\bm\Gamma|^{-1}\tr(\mathbf{R}\adj(\bm\Gamma))|_{\lambda_i = c(\mathbf{r},
    \mathbf{u})}$, which means that
  \[
    \left. \frac{\partial q(\mathbf{r} \mid \mathbf{u})}{\partial \lambda_i}
    \right|_{\lambda_i = c(\mathbf{r}, \mathbf{u})}
  \]
  has the required bounds.
\end{proof}
